---
title: "Assignment_3_Problem_2"
author: "Sai Anjesh"
date: "2024-03-12"
output: html_document
---

```{r}
## Essential modules to import
library(dplyr)
library(caret)
library(leaps)
```

### For this problem we are going to use UCI’s student performance dataset. The dataset is a recording of student grades in math and language and includes attributes related to student demographics and school related features. Click on the above link, then go to “Data Folder” and download and unzip “student.zip”. You will be using student-mat.csv file. The goal is to create a regression model to forecast student final grade in math “G3” based on the other attributes.

-----

##### 11. (0.5pt) Read the dataset into a dataframe. Ensure that you are using a correct delimiter to read the data correctly and set the “sep” option in read.csv accordingly.

```{r}
# Input data paths
student_mat_data_path <- "D:/UIS/Studies/Semester4/CSC532/Assignments/Assignment3/regression_project/input_datasets/student+performance/student/student-mat.csv"
```


```{r}
# Reading data
student_mat_data <- read.csv(file = student_mat_data_path, 
                             header = TRUE,
                             sep = ";")
```

##### 12. (2pt) Explore the dataset. More specifically, answer the following questions:

###### a. Is there any missing values in the dataset?

```{r}
# Let us check if there are any null / missing values
cat(paste(
  "Prelimiary Check for the existence of null values : ", 
  sum(is.na(student_mat_data)), 
  sep = ""))

# let us check the null values values in different formats
student_mat_data_NA_Check <- student_mat_data
student_mat_data_NA_Check[student_mat_data_NA_Check %in% c("NA", "N/A", "NULL", "", "."," ","?")] <- NA
print(sum(is.na(student_mat_data_NA_Check)))
```
So basically there are no missign values, hence we are gonna proceed further with the assignment questions.

When checked this link: https://archive.ics.uci.edu/dataset/320/student+performance
I was able to make sure that my investigation is right because the website says no maiisgin values.

###### b. Which variables are associated with the target variable G3? To answer this question, use appropriate plots and test statistics based on variable types. You can do this one by one for all variables or write a loop that applies appropriate statistic tests based on variable types. Either approach is fine.

To understand that we would first have to undertand the strucutre of the dataframe and get summary of the dataframe.

```{r}
# Structure fo the dataframe
str(student_mat_data)

# Summary
summary(student_mat_data)
```

There are 33 features. Let us check the columns and data types.

```{r}
# Code resued from my Assignemnetn 2 solution
numeric_cols <- colnames(student_mat_data)[sapply(student_mat_data, is.numeric)]
factor_cols <- colnames(student_mat_data)[sapply(student_mat_data, is.factor)]
character_cols <- colnames(student_mat_data)[sapply(student_mat_data, is.character)]

# Checking the column names
cat("Numeric columns: \n------\n", paste(numeric_cols, collapse = ", "), "\n")
cat("Factor columns: \n------\n", paste(factor_cols, collapse = ", "), "\n")
cat("Character columns: \n------\n", paste(character_cols, collapse = ", "), "\n")
```

Since we don't have factor variables hence let us make all the character variables to factor variables.

```{r}
# Let us select each character variable and make it a factor variable
student_mat_data[["school"]] <- factor(student_mat_data[["school"]], levels = c("GP","MS"))

student_mat_data[["sex"]] <- factor(student_mat_data[["sex"]], levels = c("F","M"))

student_mat_data[["address"]] <- factor(student_mat_data[["address"]], levels = c("U","R"))

student_mat_data[["famsize"]] <- factor(student_mat_data[["famsize"]], 
                                        levels = c("GT3","LE3"))

student_mat_data[["Pstatus"]] <- factor(student_mat_data[["Pstatus"]], levels = c("T","A"))

student_mat_data[["Mjob"]] <- factor(student_mat_data[["Mjob"]], 
                                     levels = c("at_home","health","other",
                                                "services","teacher"))

student_mat_data[["Fjob"]] <- factor(student_mat_data[["Fjob"]], 
                                     levels = c("teacher","other","services",
                                                "health" ,"at_home"))

student_mat_data[["reason"]] <- factor(student_mat_data[["reason"]], 
                                     levels = c("course","other","home","reputation"))

student_mat_data[["guardian"]] <- factor(student_mat_data[["guardian"]],
                                         levels = c( "mother","father","other"))

## Yes / no variables

yes_no_variables <- c("schoolsup", "famsup", "paid", 
                      "activities", "nursery", "higher", 
                      "internet", "romantic")
for(col_val in yes_no_variables) {
  student_mat_data[[col_val]] <- factor(student_mat_data[[col_val]], 
                                        levels = c( "no", "yes"))
}


```

Now let us check all the variables datatype.

```{r}
# Code resued from my Assignemnetn 2 solution
numeric_cols <- colnames(student_mat_data)[sapply(student_mat_data, is.numeric)]
factor_cols <- colnames(student_mat_data)[sapply(student_mat_data, is.factor)]
character_cols <- colnames(student_mat_data)[sapply(student_mat_data, is.character)]

# Checking the column names
cat("Numeric columns: \n------\n", paste(numeric_cols, collapse = ", "), "\n")
cat("Factor columns: \n------\n", paste(factor_cols, collapse = ", "), "\n")
cat("Character columns: \n------\n", paste(character_cols, collapse = ", "), "\n")
```

```{r}
## Reused My Code from the Assignment 1 - automation
## and made it into a function

# Nuerical vs Numerical - Correlation Test 
diagnose_association_numerical <- function(df, 
                                           numer_vars, 
                                           target_var, 
                                           signif_val = 0.05) {
  numer_p_value_dataframe <- data.frame()
  
  for(v in numer_vars[(!numer_vars %in% c(target_var))]){
    
    print(paste("*************",v,"*************",sep= ""))
    plot(x = df[, target_var], 
         y = df[,v], 
         main = paste("Scatter Plot - ",target_var," vs. ",v, sep = ""),
         xlab = target_var, 
         ylab = v)
    
    print(paste("Simple correlation check between the variables : ", cor(df[, target_var], df[, v])))
    
    # Kendall Correlation Test - to test association between 2 numerical variables
    # I am usigng the kendall rank correlation because 
    # I am comparing the G3 (score) varaible which can be ranked.
    # Also, Unlike Pearson correlation, Rank correlation can be
    # applied to both continuous (interval) and ordinal variables.
    hypotheis_test_employed <- "kendall"
    numer_Test_details <- cor.test(df[,v], df[,target_var], method = hypotheis_test_employed)
    
    print(numer_Test_details)
    
    numer_p_value_dataframe <- rbind(
      numer_p_value_dataframe,
      data.frame(
        feature1 = target_var,
        feature2 = v,
        p.value = numer_Test_details["p.value"],
        test_employed = hypotheis_test_employed
        )
      )
    }
  
  numer_p_value_dataframe$signif <- signif_val
  numer_p_value_dataframe$result <- (
    ifelse(
      numer_p_value_dataframe$p.value < numer_p_value_dataframe$signif, "pass", "fail" 
      )
    )
  
  return(numer_p_value_dataframe)
  
}

# Numerical vs Categoricl - two.test / anova test
diagnose_association_factor <- function(df,
                                        factor_vars, 
                                        target_var,
                                        signif_val = 0.05) {
  
  factor_p_value_dataframe <- data.frame()
  for(v in factor_vars[(!factor_vars %in% c(target_var))]){
    print(paste("--------------------",v,"--------------------"))
    
    # CHATGPT : how to print unique values using cat() in R in one line?
    cat("Unique values:", paste(unique(df[,v]), collapse = ", "))
    
    plot(df[, target_var]~df[,v], 
       col="green", 
       main = paste("Box Plot - ",target_var," vs. ",v, sep = ""),
       xlab = target_var,
       ylab = v)
    
    if(length(unique(df[,v])) > 2) {
      # Since we want to study the relationship between 
      # one independent categorical variable and a numeric response
      hypotheis_test_employed <- "anova"
      factor_Test_details = oneway.test(df[,target_var] ~ df[,v], 
                                        data = df)
    } else {
      # t.test - to test whether a categorical variable with two levels is
      # independent of a continuous numerical variable
      hypotheis_test_employed <- "two.test"
      factor_Test_details = t.test(df[,target_var] ~ df[,v],
                                   alternative = "two.sided")
    }
    
    factor_p_value_dataframe <- rbind(
      factor_p_value_dataframe,
      data.frame(
        feature1 = target_var,
        feature2 = v,
        p.value = factor_Test_details["p.value"],
        test_employed = hypotheis_test_employed
        )
      )
    }
  
  factor_p_value_dataframe$signif <- signif_val
  factor_p_value_dataframe$result <- (
    ifelse(
      factor_p_value_dataframe$p.value < factor_p_value_dataframe$signif,
      "pass", "fail" ))
  
  return(factor_p_value_dataframe)
}
```

```{r}
# Checking the association of variables with target variable
target_DV <- "G3"


assoc_num_df <- diagnose_association_numerical(df = student_mat_data,
                                               numer_vars = numeric_cols,
                                               target_var = target_DV)

assoc_fac_df <- diagnose_association_factor(df = student_mat_data, 
                                            factor_vars = factor_cols,
                                            target_var = target_DV)


qualified_features_data <- assoc_num_df %>% rbind.data.frame(assoc_fac_df)

qualified_features_data
```


###### c. Draw a histogram of the target variable “G3” and interpret it.

Let us draw the histogram understand the distribution.

```{r}
# Historgram of Target DV - G3
hist(student_mat_data$G3, main = "Histogram of G3", xlab = "Final Grade (G3)", ylab = "Frequency")
```
The above histogram shows us the G3 is normally distributed and 0 - 5 bin largeely concectrated towards 0 side.

#### 13. (0.5 pt) Split the data into train and test. Use 80% of samples for training and 20% of samples for testing.


```{r}
# CHATGPT: how to split dataframe into train and test

# Although the lecutre notes show splitting datasets iusing the sample function
# I still wanted to check with CHAT GPT for some other functions and ways
# I found out about caTools and other ways, train and test split used CHat CPT and I just included necessary variables

# Setting the seed for reproducibility
set.seed(123)
split_perc <- 0.8
indices <- sample(nrow(student_mat_data))
train_size <- round(split_perc * nrow(student_mat_data))

# Finding the train indices
train_indices <- indices[1:train_size]

# Create train and test dataframes
student_mat_data_train <- student_mat_data[train_indices, ]
student_mat_data_test <- student_mat_data[-train_indices, ]
```


#### 14. set the random seed: set.seed(123)

That is already set, yet doing it one more time.

```{r}
set.seed(123)
```

#### 15. (2 pt) Use caret package to run 10 fold cross validation using linear regression method on the train data to predict the “G3” variable . Print the resulting model to see the cross validation RMSE. In addition, take a summary of the model and interpret the coefficients. Which coefficients are statistically different from zero? What does this mean?

I am gonna reuse the code from the lecture notes.

```{r}
# cv Technique on train data
train.control = trainControl(method = "cv", number = 10)
model = train(G3~., data = student_mat_data_train, method = "lm", trControl = train.control)
```


Let us print the model and understand the model performance and coefficients.

```{r}
# Checking the model
print(model)

summary(model)
```

#### Set the random seed again. We need to do this before each training to ensure we get the same folds in cross validation. Set.seed(123) so we can compare the models using their cross validation RMSE.(2 pts) Use caret and leap packages to run a 10 fold cross validation using step wise linear regression method with backward selection on the train data. The train method by default uses maximum of 4 predictors and reports the best models with 1..4 predictors. We need to change this parameter to consider all predictors. So inside your train function, add the following parameter tuneGrid = data.frame(nvmax = 1:n), where n is the number of variables you use to predict “G3”.

I am gonna reuse the codes from the lecture.

```{r}
# Setting seed, althought I have set it in the previous R code chunks
set.seed(123)

# Stepwise regression
# Backwards eliminiation

number_of_predictors <- ncol(student_mat_data_train) - 1

train.control <- trainControl(method = "cv", number = 10)

step.model <- train(G3 ~., 
                    data = student_mat_data_train,
                    method = "leapBackward",
                    trControl = train.control,
                    tuneGrid = data.frame(nvmax = 1:number_of_predictors))




```

Let us check the model and its summary.

```{r}
# Checking the model
# print(step.model)

# summary(step.model)

# summary(step.model$finalModel)


# Coefficient names and values
coef_names <- names(coef(step.model$finalModel, id = 1))
coef_values <- coef(step.model$finalModel, id = 1)

# Performance metrics
rmse <- step.model$results$RMSE

# Create dataframe
model_summary <- data.frame(
  Coefficient_Names = coef_names,
  Coefficient_Values = coef_values,
  RMSE = rmse
)

# Print the dataframe
print(model_summary)


```


